Real-time resolution of short-read assembly graph using ONT long reads

### 2020-12-18 PCOMPBIOL-D-20-00393/paper_clean.tex
final change

%% END MACROS SECTION
\usepackage[normalem]{ulem}
%% npGraph
\newcommand{\npscarf}{$\mathtt{npScarf}$}
\newcommand{\npscarfg}{$\mathtt{npScarf\_wag}$}
\newcommand{\npreader}{$\mathtt{npReader}$}
\newcommand{\npanalysis}{$\mathtt{npAnalysis}$}
\newcommand{\npbarcode}{$\mathtt{npBarcode}$}
\newcommand{\npgraph}{$\mathtt{npGraph}$}
\newcommand{\canu}{$\mathtt{Canu}$}
\newcommand{\unicycler}{$\mathtt{Unicycler}$}
\newcommand{\spades}{$\mathtt{SPAdes}$}
\newcommand{\hspades}{$\mathtt{hybridSPAdes}$}
\newcommand{\albacore}{$\mathtt{Albacore}$}
\newcommand{\racon}{$\mathtt{Racon}$}
\newcommand{\metrichor}{$\mathtt{Metrichor}$}
\newcommand{\minimap}{$\mathtt{minimap2}$}
\newcommand{\miniasm}{$\mathtt{miniasm}$}
\newcommand{\bwa}{$\mathtt{BWA\text{-}MEM}$}
\newcommand{\lrscaf}{$\mathtt{LRScaf}$}

@@

\newcommand{\IE}{\emph{i.e.}}
\newcommand{\EG}{\emph{e.g.}}
\newcommand{\review}[1]{\textcolor{red}{#1}}

@@

%% End preloads

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Real-time resolution of short-read assembly graph using ONT long reads} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Son Hoang Nguyen\textsuperscript{1*},
Minh Duc Cao\textsuperscript{1},
Lachlan Coin\textsuperscript{1,2,3,4*}
\\
\bigskip
\textbf{1} Institute for Molecular Bioscience, the University of Queensland, 
St Lucia, Brisbane, QLD 4072 Australia
\\
\textbf{2} Department of Microbiology and Immunology, The University of Melbourne, Parkville, 3010, Australia
\\
\textbf{3} Department of Clinical Pathology, The University of Melbourne, Parkville, 3010, Australia
\\
\textbf{4} Department of Infectious Disease, Imperial College London, London, W2 1NY, UK
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
*s.hoangnguyen@imb.uq.edu.au,l.coin@imb.uq.edu.au

@@

\npgraph{} aligns long reads to the contigs in the assembly graph. When a long read is aligned
to multiple contigs, \npgraph{} constructs candidate paths that are supported by the read. This strategy allows \npgraph{} to progressively update the likelihood of the paths going through a knot.
When sufficient data are obtained, the best path is confidently identified and hence the knot is untangled.
%TODO: two more sentences describing the bridging
In general, the bridging algorithm used to determine the best path is a combination of progressive merging, accumulated scoring and decision making modules.
It operates on each pair of unique contigs, or \emph{anchors}, by using a \emph{bridge} data structure maintaining 2 anchors and a list of \emph{steps} in-between as shown in Figure~\ref{figure:npgraph_pipeline}(a). 
A set of candidate paths are listed and the best one can be selected amongst them given enough evidence. 
The de Bruijin graph is subsequently simplified when the bridge is replaced by a single edge representing the best path (\ref{figure:npgraph_pipeline}(b)).

We also provide a Graphical User Interface (GUI) for \npgraph{}. The GUI includes the dashboard for controlling the settings of the program and a window for visualization of the assembly graph 
in real-time (Figure~\ref{figure:npgraph_gui}). 
In this interface, the assembly graph loading stage is separated from the actual assembly process so that users can check for the graph quality first before carry out any further tasks.
A proper combination of command line and GUI can provide an useful streaming pipeline that copes well with MinION output data. This is designed to support the real-time monitoring of
the results from real-time sequencing~\cite{CaoGC2016,Cao2017scaffolding,Nguyen2017barcode} that allow the analysis to take place abreast to a nanopore sequencing run.

@@

At early stages (level 1 or 2), a bridge is constructed progressively by alignments from long reads that spanning its corresponding anchor(s).
In an example from Figure~\ref{figure:npgraph_pipeline}(a), bridges from a certain anchor (highlighted in red) are created by extracting appropriate alignments from incoming long reads to the contigs. Each of the steps therefore is assigned a weighing score based on its alignment quality.
Due to the error rate of long reads, there should be deviations in terms of steps found and spans measured between these bridges, even though they represent the same connection.
A continuous merging phase, as shown in the figure, takes advantage of a pairwise Needleman-Wunsch dynamic programming to generate a consensus list based on weight and position of each of every stepping nodes. The spans are calibrated accordingly by averaging out the distances. On the other hand, the score of the merged steps are accumulated over time as well.
Whenever a consensus bridge is anchored by 2 unique contigs at both ends and hosting a list of steps with sufficient coverage, it is ready for a path finding in the next step.

@@

Figure~\ref{figure:npgraph_graphview} presents an example of the results before and after graph resolving process in the GUI.
The result graph, after cleaning, would only report the significant connected components that represents the final contigs.
Smaller fragments, even unfinished but with high remaining coverage, are also presented as potential candidates for further downstream analysis.
Further annotation utility can be implemented in the future better monitoring the features of interests as in \npscarf{}.
\subsubsection*{Result extraction and output}
\npgraph{} reports assembly result in real-time by decomposing the assembly graph into a set of longest straight paths (LSP), each of the LSP will spell a contig in the assembly report.
The final assembly output contains files in both FASTA and GFAv1 format (\url{https://github.com/GFA-spec/GFA-spec}). While the former only retains the actual genome sequences from the final decomposed graph, the latter output file can store almost every properties of the ultimate graph such as nodes, links and potential paths between them.

A path $p=\{v_0,e_1,v_1,\ldots,v_{k-1},e_k,v_k\}$ of size $k$ is considered as straight if and only if each of every edges along the path $e_i, \forall i=1,\ldots,k$ is the only option to traverse from either $v_{i-1}$ or $v_i$, giving the transition rule.
To decompose the graph, the tool simply mask out all incoming/outgoing edges rooted from any node with in/out degree greater than 1 as demonstrated in Figure~\ref{figure:npgraph_decompose}. These edges are defined as branching edges which stop straight paths from further extending.

@@

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Evaluation using synthetic data}
To evaluate the performance of the method, \npgraph{} was tested along with \spades{}, \spades{} hybrid from version 3.13.1, ~\cite{AntipovKM2015}, \npscarf{} (japsa 1.7-02), \lrscaf{} version 1.1.9 and \unicycler{} version 0.4.6 using \unicycler{}`s synthetic data set~ \cite{Wick2017unicycler} . 
The data set is a simulation of Illumina and MinION raw data, generated \emph{in silico} based on available microbial references.
We ran hybrid assembly methods using the entire nanopore data and the reciprocal results were evaluated by QUAST 5.0.2~\cite{Mikheenko2018quast5}.

@@

To align the long reads to the assembly graph components, both \bwa{}~\cite{Li2013} or \minimap{}~\cite{Li2016} were used in conjunction with \npgraph{}. These two methods were chosen due to their proven efficiency and compatibility with streaming data.
While \bwa{} is a well-known classic aligner that can be adapted to work with third generation sequencing data, \minimap{} has been specially designed for this data type. 
We observed a slightly higher error rate (comprising the sum of mismatches and indels per 100kb) using \bwa{} in comparison to \minimap{} for all simulations in Table~\ref{table:npgraph_compare}.  This is due to the fact that bridging paths induced using \bwa{} were slightly less accurate due to more noise from the smaller \emph{steps} in-between (Figure~\ref{figure:npgraph_pipeline}(a)).
However, under almost circumstances, using either aligner resulted in final assemblies with comparable qualities.
In terms of running time and resources required, \minimap{} always proved to be a better option, requiring markedly less CPU time than \bwa{}.  Utilising \minimap, \npgraph{} is now the fastest hybrid assembler available.

@@

In addition to what is presented in Table~\ref{tab:npgraph_real}, dot plots for the pair-wise alignments between the assembly contigs were generated and can be found in Figure S1. This identified a structural difference between \npgraph{} and \unicycler{} assembly for the \emph{E.~cloacae} CAV1411 genome assembly. This was caused by the inconsistency of a fragment's direction on the final output contigs. Comparison to a reference 
 genome from the same bacteria strain (GenBank ID: CP011581.1~\cite{Potter2016rapid}), demonstrated that contigs generated by \npgraph{} produced consistent alignment, but not those generated by \unicycler{} (Figure S2). However, we cannot at this stage rule out genuine structural variation between the two samples .

\subsection*{Assembly performance on streaming data}
In order to investigate the rate at which the two streaming hybrid assembly algorithms completed bacterial assemblies, we plot the N50 as a function of long-read coverage on the 4 datasets described in the previous section ( Figure~\ref{F:npgraph_rt}) .This revealed that \npgraph{} and \npscarf{} both converge to the same ultimate completeness but at different rates. \npscarf{} converged more quickly than \npgraph{}, due to the fact that it is able to build bridges with only 1 spanning long-read, whereas \npgraph{} requires 3 reads.    
Unlike \npscarf{} where the connections could be undone and rectified later if needed, a bridge in \npgraph{} will remain unchanged once created.
The plot for \ec{} data clarifies this behaviour when a fluctuation can be observed in \npscarf{} assembly at $\simeq 3$-folds data coverage.
On the other hand, the N50 length of \npgraph{} is always a monotonic increasing function. 
The sharp \emph{jumping} patterns suggested that the linking information from long-read data had been stored and exploited at certain time point decided by the algorithm. 
In addition, at the end of the streaming when the sequencing is finished, \npgraph{} will try for the last time to connect bridges with less than 3 supporting reads which are otherwise not part of conflicting bridges.

@@

\section*{Data availability}
All data used for benchmarking were from public sources. The synthetic data was downloaded from \url{https://cloudstor.aarnet.edu.au/plus/index.php/s/dzRCaxLjpGpfKYW}. The real bacterial sequencing data was from NCBI under project accession number PRJNA353060.

\section*{Acknowledgments}
This work was supported by an ARC grant to LC and MDC (DP170102626).  LC is funded by NHMRC CDF (GNT1130084).  This research was supported by the use of the NeCTAR Research Cloud, by QCIF and by the University of Queensland’s Research Computing Centre (RCC).

@@

\paragraph*{Fig S1}
Dotplot generated by MUMmer for assembly results of \unicycler{} versus \npgraph{}. Structural agreements between two methods were found in (a)~\emph{C.freundii} and (b)~\emph{K.oxytoca} assembly contigs. On the other hand, for (c)~\emph{E.cloacae} sample, there was a disagreement detected between 2 largest contigs given by two assembly algorithms.

\paragraph*{Fig S2}
Alignments of an \emph{Enterobacter~cloacae} reference genome to assembly sequences generated by  (a)~\npgraph{} and (b)~\unicycler{}. The former suggests a structural variant, the latter is virtually an 1-to-1 mapping.

\paragraph*{Table S1}
Benchmarking different methods using \lrscaf{}, \npscarf{}, \npgraph{}, $\mathtt{hybridSPAdes}$ and \unicycler{} hybrid assembler with the synthetic data set.

\paragraph*{Fig5.csv}
Spreadsheet contains data used to generate Fig 5.

>>>>

%% END MACROS SECTION
\usepackage[normalem]{ulem}

%% npGraph
\newcommand{\npscarf}{$\mathtt{npScarf}$}
\newcommand{\npscarfg}{$\mathtt{npScarf\_wag}$}
\newcommand{\npreader}{$\mathtt{npReader}$}
\newcommand{\npanalysis}{$\mathtt{npAnalysis}$}
\newcommand{\npbarcode}{$\mathtt{npBarcode}$}
\newcommand{\npgraph}{$\mathtt{npGraph}$}
\newcommand{\canu}{$\mathtt{Canu}$}
\newcommand{\unicycler}{$\mathtt{Unicycler}$}
\newcommand{\spades}{$\mathtt{SPAdes}$}
\newcommand{\hspades}{$\mathtt{hybridSPAdes}$}
\newcommand{\albacore}{$\mathtt{Albacore}$}
\newcommand{\racon}{$\mathtt{Racon}$}
\newcommand{\metrichor}{$\mathtt{Metrichor}$}
\newcommand{\minimap}{$\mathtt{minimap2}$}
\newcommand{\miniasm}{$\mathtt{miniasm}$}
\newcommand{\bwa}{$\mathtt{BWA\text{-}MEM}$}
\newcommand{\lrscaf}{$\mathtt{LRScaf}$}

@@

\newcommand{\IE}{\emph{i.e.}}
\newcommand{\EG}{\emph{e.g.}}

@@

%% For tracked changes
\usepackage{changes}
\definechangesauthor[name={Son Nguyen}, color=red]{SN}
\colorlet{Changes@Color}{red}
%% End preloads

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Real-time resolution of short-read assembly graph using ONT long reads} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Son Hoang Nguyen\textsuperscript{1*},
Minh Duc Cao\textsuperscript{1},
Lachlan Coin\textsuperscript{1,2,3,4*}
\\
\bigskip
\textbf{1} Institute for Molecular Bioscience, the University of Queensland, 
St Lucia, Brisbane, Australia
\\
\textbf{2} Department of Microbiology and Immunology, The University of Melbourne, Parkville, Australia
\\
\textbf{3} Department of Clinical Pathology, The University of Melbourne, Parkville, Australia
\\
\textbf{4} Department of Infectious Disease, Imperial College London, London, UK
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
*SHN: s.hoangnguyen@imb.uq.edu.au, LC: l.coin@imb.uq.edu.au

@@

\npgraph{} aligns long reads to the contigs in the assembly graph. When a long read is aligned
to multiple contigs, \npgraph{} constructs candidate paths that are supported by the read. This strategy allows \npgraph{} to progressively update the likelihood of the paths going through a knot.
When sufficient data are obtained, the best path is confidently identified and hence the knot is untangled.
%TODO: two more sentences describing the bridging
In general, the bridging algorithm used to determine the best path is a combination of progressive merging, accumulated scoring and decision making modules.
It operates on each pair of unique contigs, or \emph{anchors}, by using a \emph{bridge} data structure maintaining 2 anchors and a list of \emph{steps} in-between as shown in Fig~\ref{figure:npgraph_pipeline}A. 
A set of candidate paths are listed and the best one can be selected amongst them given enough evidence. 
The de Bruijin graph is subsequently simplified when the bridge is replaced by a single edge representing the best path (Fig~\ref{figure:npgraph_pipeline}B).

We also provide a Graphical User Interface (GUI) for \npgraph{}. The GUI includes the dashboard for controlling the settings of the program and a window for visualization of the assembly graph 
in real-time (Fig~\ref{figure:npgraph_gui}). 
In this interface, the assembly graph loading stage is separated from the actual assembly process so that users can check for the graph quality first before carry out any further tasks.
A proper combination of command line and GUI can provide an useful streaming pipeline that copes well with MinION output data. This is designed to support the real-time monitoring of
the results from real-time sequencing~\cite{CaoGC2016,Cao2017scaffolding,Nguyen2017barcode} that allow the analysis to take place abreast to a nanopore sequencing run.

@@

At early stages (level 1 or 2), a bridge is constructed progressively by alignments from long reads that spanning its corresponding anchor(s).
In an example from Fig~\ref{figure:npgraph_pipeline}A, bridges from a certain anchor (highlighted in red) are created by extracting appropriate alignments from incoming long reads to the contigs. Each of the steps therefore is assigned a weighing score based on its alignment quality.
Due to the error rate of long reads, there should be deviations in terms of steps found and spans measured between these bridges, even though they represent the same connection.
A continuous merging phase, as shown in the figure, takes advantage of a pairwise Needleman-Wunsch dynamic programming to generate a consensus list based on weight and position of each of every stepping nodes. The spans are calibrated accordingly by averaging out the distances. On the other hand, the score of the merged steps are accumulated over time as well.
Whenever a consensus bridge is anchored by 2 unique contigs at both ends and hosting a list of steps with sufficient coverage, it is ready for a path finding in the next step.

@@

Fig~\ref{figure:npgraph_graphview} presents an example of the results before and after graph resolving process in the GUI.
The result graph, after cleaning, would only report the significant connected components that represents the final contigs.
Smaller fragments, even unfinished but with high remaining coverage, are also presented as potential candidates for further downstream analysis.
Further annotation utility can be implemented in the future better monitoring the features of interests as in \npscarf{}.
\subsubsection*{Result extraction and output}
\npgraph{} reports assembly result in real-time by decomposing the assembly graph into a set of longest straight paths (LSP), each of the LSP will spell a contig in the assembly report.
The final assembly output contains files in both FASTA and GFAv1 format (\url{https://github.com/GFA-spec/GFA-spec}). While the former only retains the actual genome sequences from the final decomposed graph, the latter output file can store almost every properties of the ultimate graph such as nodes, links and potential paths between them.

A path $p=\{v_0,e_1,v_1,\ldots,v_{k-1},e_k,v_k\}$ of size $k$ is considered as straight if and only if each of every edges along the path $e_i, \forall i=1,\ldots,k$ is the only option to traverse from either $v_{i-1}$ or $v_i$, giving the transition rule.
To decompose the graph, the tool simply mask out all incoming/outgoing edges rooted from any node with in/out degree greater than 1 as demonstrated in Fig~\ref{figure:npgraph_decompose}. These edges are defined as branching edges which stop straight paths from further extending.

@@

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Evaluation using synthetic data}
To evaluate the performance of the method, \npgraph{} was tested along with \spades{}, \spades{} hybrid from version 3.13.1, ~\cite{AntipovKM2015}, \npscarf{} (japsa 1.7-02), \lrscaf{} version 1.1.9~\cite{Qin2019} and \unicycler{} version 0.4.6 using \unicycler{}`s synthetic data set~\cite{Wick2017unicycler} . 
The data set is a simulation of Illumina and MinION raw data, generated \emph{in silico} based on available microbial references.
We ran hybrid assembly methods using the entire nanopore data and the reciprocal results were evaluated by QUAST 5.0.2~\cite{Mikheenko2018quast5}.

@@

To align the long reads to the assembly graph components, both \bwa{}~\cite{Li2013} or \minimap{}~\cite{Li2016} were used in conjunction with \npgraph{}. These two methods were chosen due to their proven efficiency and compatibility with streaming data.
While \bwa{} is a well-known classic aligner that can be adapted to work with third generation sequencing data, \minimap{} has been specially designed for this data type. 
We observed a slightly higher error rate (comprising the sum of mismatches and indels per 100kb) using \bwa{} in comparison to \minimap{} for all simulations in Table~\ref{table:npgraph_compare}.  This is due to the fact that bridging paths induced using \bwa{} were slightly less accurate due to more noise from the smaller \emph{steps} in-between (Fig~\ref{figure:npgraph_pipeline}A).
However, under almost circumstances, using either aligner resulted in final assemblies with comparable qualities.
In terms of running time and resources required, \minimap{} always proved to be a better option, requiring markedly less CPU time than \bwa{}.  Utilising \minimap, \npgraph{} is now the fastest hybrid assembler available.

@@

In addition to what is presented in Table~\ref{tab:npgraph_real}, dot plots for the pair-wise alignments between the assembly contigs were generated and can be found in S1 Fig. This identified a structural difference between \npgraph{} and \unicycler{} assembly for the \emph{E.~cloacae} CAV1411 genome assembly. This was caused by the inconsistency of a fragment's direction on the final output contigs. Comparison to a reference 
 genome from the same bacteria strain (GenBank ID: CP011581.1~\cite{Potter2016rapid}), demonstrated that contigs generated by \npgraph{} produced consistent alignment, but not those generated by \unicycler{} (S2 Fig). However, we cannot at this stage rule out genuine structural variation between the two samples .

\subsection*{Assembly performance on streaming data}
In order to investigate the rate at which the two streaming hybrid assembly algorithms completed bacterial assemblies, we plot the N50 as a function of long-read coverage on the 4 datasets described in the previous section ( Fig~\ref{F:npgraph_rt}).
This revealed that \npgraph{} and \npscarf{} both converge to the same ultimate completeness but at different rates. \npscarf{} converged more quickly than \npgraph{}, due to the fact that it is able to build bridges with only 1 spanning long-read, whereas \npgraph{} requires 3 reads.    
Unlike \npscarf{} where the connections could be undone and rectified later if needed, a bridge in \npgraph{} will remain unchanged once created.
The plot for \ec{} data clarifies this behaviour when a fluctuation can be observed in \npscarf{} assembly at $\simeq 3$-folds data coverage.
On the other hand, the N50 length of \npgraph{} is always a monotonic increasing function. 
The sharp \emph{jumping} patterns suggested that the linking information from long-read data had been stored and exploited at certain time point decided by the algorithm. 
In addition, at the end of the streaming when the sequencing is finished, \npgraph{} will try for the last time to connect bridges with less than 3 supporting reads which are otherwise not part of conflicting bridges.
S1 Data provides the results in detail and were used to generate Fig~\ref{F:npgraph_rt}.

@@

\section*{Acknowledgments}
This research was supported by the use of the NeCTAR Research Cloud, by QCIF and by the University of Queensland’s Research Computing Centre (RCC).

@@

\paragraph*{S1 Fig}
Dotplot generated by MUMmer for assembly results of \unicycler{} versus \npgraph{}. Structural agreements between two methods were found in (a)~\emph{C.freundii} and (b)~\emph{K.oxytoca} assembly contigs. On the other hand, for (c)~\emph{E.cloacae} sample, there was a disagreement detected between 2 largest contigs given by two assembly algorithms.

\paragraph*{S2 Fig}
Alignments of an \emph{Enterobacter~cloacae} reference genome to assembly sequences generated by  (a)~\npgraph{} and (b)~\unicycler{}. The former suggests a structural variant, the latter is virtually an 1-to-1 mapping.

\paragraph*{S1 Table}
Benchmarking different methods using \lrscaf{}, \npscarf{}, \npgraph{}, $\mathtt{hybridSPAdes}$ and \unicycler{} hybrid assembler with the synthetic data set.

\paragraph*{S1 Data}
Spreadsheet contains data used to generate Fig 5.
